Chemistry of the Orphan Stream I: Identifying Stream Members from Low-Resolution Spectroscopy

### 2012-09-16 2012-oss-mrs/draft-oss-mrs.tex
All corrections made

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Anglo-Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.71$ with an intrinsic dispersion of $\sigma = 0.56$\,dex which is more metal-rich than previously found and unbiased by spectral type. The metallicity dispersion in our stream sample is indicative that the Orphan Stream has a dSph origin. We find a revised stream distance of $23.5 \pm 1.0$\,kpc near the celestial equator, and our kinematic signature peaks at $V_{\textsc{GSR}} \sim ZZ$\,km s$^{-1}$. The observed velocity dispersion of our highest probability members is consistent with arising from the velocity uncertainties alone, indicating that, at least along the line-of-sight, the Orphan Stream is kinematically quite cold. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of ten highly probable Orphan Stream candidates, all worthy of high-resolution spectroscopic follow up. In the following section we outline our photometric target selection. In \S\ref{sec:observations} we describe the low-resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.

@@

\subsection{Kinematics}
Radial velocities were measured by cross-correlating our normalised spectra against a K-giant synthetic template with a temperature of 4500\,K, $\log{g}$ = 1 and $[\mbox{Fe/H}] = -1.5$ across the range $845 \leq \lambda \leq 870$\,nm. Heliocentric velocities were translated to the galactic rest frame by adopting the local standard of rest velocity as 220\,km s$^{-1}$ towards $(l, b) = (53\,^\circ, 25\,^\circ)$ \citep{Kerr;Lynden-Bell_1986, Mihalas;Binney_1981}\footnote{Where $V_{\textsc{gsr}} = V_{\textsc{helio}} + 220\sin{l}\cos{b} + 16.5\times[\sin{b}\sin{25} + \cos{b}\cos{25}\cos{(l - 53)}]$}. \citet{Newberg;et-al_2010} presented a comprehensive kinematic dataset on the Orphan Stream using BHB stars and employed a slightly different galactocentric transformation. This systematic difference equates to our galactocentric rest frame velocities being on average $\sim5$\,km s$^{-1}$ lower than those of \citet{Newberg;et-al_2010}.

Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{\textsc{gsr}} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no sharp kinematic peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

In this region \citet{Newberg;et-al_2010} found the Orphan Stream to have $V_{\textsc{gsr}} = 110$ km s$^{-1}$ from BHB stars, approximately $105$\,km s$^{-1}$ on our scale. To isolate potential Orphan Stream members we have nominated a relatively wide kinematic selection region between $75 \leq V_{\textsc{gsr}} \leq 135$ km s$^{-1}$, which yields 25 Orphan Stream candidates. The typical uncertainty in our kinematics is $\pm{}3.5$\,km s$^{-1}$, although for our fainter targets this quickly increases above 10\,km s$^{-1}$.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff;Da_Costa_1991}. We have employed the \citet{Starkenburg;et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg;et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane;et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/feh.pdf}
	\caption{Metallicities from the Ca \textsc{II} triplet lines against metallicities derived from isochrone fitting for the eighteen stars which meet our kinematic and surface gravity criteria. Both abundance determinations imply these stars are RGB members of the Orphan Stream at a distance of $\sim21.4$\,kpc \citep{Newberg;et-al_2010}. Consistency between these methods indicates highly likely stream membership (green region). The minimum isochrone metallicity, and a representative uncertainty of 0.2\,dex for abundance measurements is shown.}
	\label{fig:feh}
\end{figure}

Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10 Gyr \citet{Girardi;et-al_2008} isochrone at 21.4\,kpc \citep{Newberg;et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone (see Figure \ref{fig:feh}). Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting which are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim21.4$\,kpc. We find ten highly likely stream members with consistently derived metallicities, and they fall within the green shaded region illustrated in Figure \ref{fig:feh}.

@@

@@

\begin{deluxetable*}{lcccccccccccc}
\tabletypesize{\scriptsize}
\tablecaption{Identified Orphan Stream Members\label{tab:oss-members}}
\tablehead{
	\colhead{$\alpha$} &
	\colhead{$\delta$} &
	\colhead{$g$} &
	\colhead{$g \-- r$} &
	\colhead{$V_{hel}$} &
	\colhead{$V_{\textsc{gsr}}$} &
	\colhead{$V_{err}$} &
	\colhead{$EW_{\lambda8807}$} &
	\colhead{[Fe/H]$_{Ca}$} &
	\colhead{[Fe/H]$_{iso}$} &
	\colhead{[Fe/H]\tablenotemark{a}} &
	\colhead{Stream} \\
(J2000) & (J2000) & & & (kms $^{-1}$) & (km s$^{-1}$) & (km s$^{-1}$) & (\AA) & (dex) & (dex) & (dex) & Prob.
}
\startdata
10 46 29.3 & $-$00 19 38.5 & 16.97 & 0.56 & ? &  78.4 &  5.2 &    0.126 & --1.63  &    --1.68 & --1.63 & High \\
10 47 06.1 & $-$01 56 03.9 & 17.98 & 0.54 & ? &  74.9 & 17.6 &    0.452 & :--1.12\tablenotemark{b} &    --1.40 & --1.40 & High \\
10 47 17.6 & $+$00 25 07.7 & 14.89 & 0.73 & ? &  79.2 &  3.0 &    0.213 & --1.77  &    --1.80 & --1.77 & High \\
10 47 29.1 & $-$02 02 22.6 & 17.20 & 0.46 & ? &  93.2 & 29.8 & $<$0.22  & --2.82  & $<$--2.28 & --2.82 & High \\
10 47 30.1 & $-$00 01 24.6 & 17.25 & 0.61 & ? &  83.5 &  3.2 &    0.123 & --1.62  &    --1.68 & --1.68 & High \\
10 48 20.9 & $+$00 26 34.4 & 17.88 & 0.55 & ? & 118.8 & 11.7 &    0.467 & --1.65  &    --1.73 & --1.73 & High \\
10 48 44.4 & $-$02 53 08.8 & 18.35 & 0.62 & ? & 108.1 &  9.0 &    0.183 & :--1.01\tablenotemark{b} &    --1.17 & --1.17 & High \\
10 48 46.9 & $-$00 32 27.8 & 17.85 & 0.46 & ? & 109.3 &  8.1 &    0.324 & --2.37  & $<$--2.28 & --2.37 & High \\
10 49 08.3 & $+$00 02 00.2 & 16.27 & 0.62 & ? &  81.5 &  4.6 &    0.034 & --2.70  & $<$--2.28 & --2.70 & High \\
10 51 19.7 & $+$00 05 15.4 & 17.81 & 0.65 & ? &  82.6 &  5.0 &    0.198 & --1.16  &    --1.20 & --1.20 & High 
\enddata
\tablenotetext{a}{Final adopted [Fe/H] value based on quality of our derived metallicities.}
\tablenotetext{b}{Sufficiently fainter than $g_{HB}$ to qualify this measurement as uncertain.}
\end{deluxetable*}

@@

Using a maximum-likelihood estimation we find the mean stream galactocentric velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = ZZ.Z \pm Z.Z$\,km s$^{-1}$ and the intrinsic dispersion to be $Z.Z \pm Z.Z$\,km s$^{-1}$, i.e., comparable to our velocity resolution and consistent with zero. If we exclude five stars with weak signal \--- and hence large velocity uncertainties \--- this dispersion is found to be $Z \pm Z$\,km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.

The K giants presented here can provide great insight into the chemistry, and history of the Orphan Stream. High-resolution spectroscopic observations have been taken for our most highly probable members and a detailed chemical analysis will be presented in a forthcoming paper (Casey et al. 2012c, in preparation). Detailed chemical abundances can help determine both the nature of the host system, and allow us to compare peculiar chemical signatures with those of the known Milky Way satellites in order to associate likely parents. However, at least for the moment, the Orphan Stream remains appropriately named.

>>>>

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Anglo-Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.71$ with an intrinsic dispersion of $\sigma = 0.56$\,dex which is more metal-rich than previously found and unbiased by spectral type. The metallicity dispersion in our stream sample is indicative that the Orphan Stream has a dSph origin. We find a revised stream distance of $23.5 \pm 1.0$\,kpc near the celestial equator, and our kinematic signature peaks at $V_{GSR} = 00.0$\,km s$^{-1}$. The observed velocity dispersion of our highest probability members is consistent with arising from the velocity uncertainties alone, indicating that, at least along the line-of-sight, the Orphan Stream is kinematically quite cold. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of ten highly probable Orphan Stream candidates, all worthy of high-resolution spectroscopic follow up. In the following section we outline our photometric target selection. In \S\ref{sec:observations} we describe the low-resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.
\newline

@@

\subsection{Kinematics}
Radial velocities were measured by cross-correlating our normalised spectra against a K-giant synthetic template with a temperature of 4500\,K, $\log{g}$ = 1 and $[\mbox{Fe/H}] = -1.5$ across the range $845 \leq \lambda \leq 870$\,nm. Heliocentric velocities were translated to the galactic rest frame by adopting the local standard of rest velocity as 220\,km s$^{-1}$ towards $(l, b) = (53\,^\circ, 25\,^\circ)$ \citep{Kerr;Lynden-Bell_1986, Mihalas;Binney_1981}\footnote{Where $V_{GSR} = V_{\textsc{helio}} + 220\sin{l}\cos{b} + 16.5\times[\sin{b}\sin{25} + \cos{b}\cos{25}\cos{(l - 53)}]$}. \citet{Newberg;et-al_2010} presented a comprehensive kinematic dataset on the Orphan Stream using BHB stars and employed a slightly different galactocentric transformation. This systematic difference equates to our galactocentric rest frame velocities being on average $\sim5$\,km s$^{-1}$ lower than those of \citet{Newberg;et-al_2010}.

@@

Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. We have selected predicted stars from the Besan\c{c}on model using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{GSR} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. There is no sharp kinematic peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

In this region \citet{Newberg;et-al_2010} found the Orphan Stream to have $V_{GSR} = 110$ km s$^{-1}$ from BHB stars, approximately $105$\,km s$^{-1}$ on our scale. To isolate potential Orphan Stream members we have nominated a relatively wide kinematic selection region between $75 \leq V_{GSR} \leq 135$ km s$^{-1}$, which yields 25 Orphan Stream candidates. The typical uncertainty in our kinematics is $\pm{}3.5$\,km s$^{-1}$, although for our fainter targets this quickly increases above 10\,km s$^{-1}$.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff;Da_Costa_1991}. We have employed the \citet{Starkenburg;et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg;et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane;et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.
Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10 Gyr \citet{Girardi;et-al_2008} isochrone at 21.4\,kpc \citep{Newberg;et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone (see Figure \ref{fig:feh}). Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting that are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim21.4$\,kpc. We find ten highly likely stream members with consistently derived metallicities. They fall within the shaded region illustrated in Figure \ref{fig:feh}.

@@

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/feh.pdf}
	\caption{Metallicities from the Ca\,\textsc{II} triplet lines versus those found from fitting isochrones to the eighteen stars that meet our kinematic and surface gravity criteria. Both abundance determinations imply these stars are RGB members of the Orphan Stream at a distance of $\sim$21.4\,kpc \citep{Newberg;et-al_2010}. Consistency between these methods indicates highly likely stream membership (green region). The minimum isochrone metallicity, and a representative uncertainty of 0.2\,dex for abundance measurements is shown.}
	\label{fig:feh}
\end{figure}

@@

\begin{deluxetable*}{ccccccccccccc}
\tabletypesize{\scriptsize}
\tablecaption{Identified Orphan Stream Candidates\label{tab:oss-members}}
\tablehead{
	\colhead{$\alpha$} &
	\colhead{$\delta$} &
	\colhead{$g$} &
	\colhead{$g \-- r$} &
	\colhead{$V_{hel}$} &
	\colhead{$V_{GSR}$} &
	\colhead{$V_{err}$} &
	\colhead{$EW_{\lambda8807}$} &
	\colhead{[Fe/H]$_{Ca}$} &
	\colhead{[Fe/H]$_{iso}$} &
	\colhead{[Fe/H]\tablenotemark{a}} &
	\colhead{Stream} \\
(J2000) & (J2000) & & & (kms $^{-1}$) & (km s$^{-1}$) & (km s$^{-1}$) & (\AA) & (dex) & (dex) & (dex) & Prob.
}
\startdata
10:46:21.9 & $+$00:43:21.8 & 17.52 & 0.50 & 209.2 &  73.3 &  9.3 & 0.273 & --1.78 &$<$--2.28& --1.78 & Low \\
10:46:29.3 & $-$00:19:38.5 & 17.77 & 0.56 & 217.4 &  78.4 &  5.2 & 0.126 & --1.63 & --1.68  & --1.63 & High \\
10:46:50.4 & $-$00:13:15.6 & 17.33 & 0.51 & 215.5 &  77.0 &  4.0 & 0.416 & --1.31 &$<$--2.28& --1.31 & Low \\
10:47:06.1 & $-$01:56:03.9 & 18.74 & 0.54 & 218.4 &  74.9 & 17.6 & 0.452 &:--1.12\tablenotemark{b}& --1.40  & --1.40 & High \\
10:47:15.0 & $-$03:15:03.9 & 18.66 & 0.54 & 256.8 & 109.5 &  9.0 &$<$0.19&:--1.85\tablenotemark{b}& --1.43  & --1.43 & Medium \\
10:47:17.6 & $+$00:25:07.7 & 16.09 & 0.72 & 215.7 &  79.2 &  3.3 & 0.212 & --1.84 & --1.80  & --1.84 & High \\
10:47:29.1 & $-$02:02:22.6 & 17.86 & 0.47 & 236.9 &  93.2 & 29.8 &$<$0.40& --2.82 &$<$--2.28& --2.82 & High \\
10:47:30.1 & $-$00:01:24.5 & 17.25 & 0.61 & 221.3 &  83.6 &  3.5 & 0.123 & --1.62 & --1.68  & --1.62 & High \\
10:48:20.9 & $+$00:26:34.4 & 17.88 & 0.55 & 254.9 & 118.9 & 11.7 & 0.467 & --1.65 & --1.73  & --1.65 & High \\
10:48:27.8 & $+$00:55:24.0 & 17.72 & 0.51 & 259.1 & 124.5 &  6.7 & 0.182 & --1.48 &$<$--2.28& --1.48 & Low \\
10:48:31.9 & $+$00:03:35.7 & 17.02 & 0.58 & 242.3 & 105.1 &  5.1 & 0.234 & --1.12 & --2.10  & --1.12 & Low \\
10:48:44.4 & $-$02:53:08.8 & 18.35 & 0.62 & 253.9 & 108.2 &  9.0 & 0.183 &:--1.01\tablenotemark{b}& --1.17  & --1.17 & High \\
10:48:46.9 & $-$00:32:27.8 & 17.85 & 0.46 & 248.2 & 109.3 &  8.1 & 0.324 & --2.37 &$<$--2.28& --2.37 & High \\
10:49:08.3 & $+$00:02:00.2 & 16.27 & 0.62 & 218.5 &  81.5 &  4.6 & 0.034 & --2.70 &$<$--2.28& --2.70 & High \\
10:50:13.1 & $+$00:33:52.7 & 16.13 & 0.58 & 229.7 &  94.7 &  5.1 & 0.391 & --1.54 &$<$--2.28& --1.54 & Low \\
10:50:24.2 & $-$01:49:05.4 & 17.94 & 0.54 & 252.0 & 109.9 & 25.4 & 0.151 & --1.06 & --1.73  & --1.06 & Low \\
10:50:33.8 & $+$00:12:19.1 & 17.82 & 0.60 & 233.5 &  97.5 &  5.9 & 0.596 & --0.90 & --1.43  & --0.90 & Medium \\
10:51:19.7 & $+$00:05:15.5 & 17.81 & 0.65 & 218.7 &  82.7 &  5.0 & 0.198 & --1.16 & --1.20  & --1.16 & High 
\enddata
\tablenotetext{a}{Final adopted [Fe/H] value based on quality of two metallicity measurements.}
\tablenotetext{b}{Sufficiently fainter than $g_{HB}$ to qualify this measurement as uncertain.}
\end{deluxetable*}

@@

Using a maximum-likelihood estimation we find the stream velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = 00.0 \pm 0.0$\,km s$^{-1}$ and the intrinsic dispersion to be $0.0 \pm 0.0$\,km s$^{-1}$, i.e., comparable to our velocity resolution and consistent with zero. If we exclude five stars with weak signal \--- and hence large velocity uncertainties \--- this dispersion is found to be $0.0 \pm 0.0$\,km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.

The K giants presented here can provide great insight into the chemistry, and history of the Orphan Stream. High-resolution spectroscopic observations have been taken for our most highly probable members and a detailed chemical analysis will be presented in a forthcoming paper (Casey et al. 2012c, in prep.). Detailed chemical abundances can help determine both the nature of the host system, and allow us to compare peculiar chemical signatures with those of the known Milky Way satellites in order to associate likely parents. However, at least for the moment, the Orphan Stream remains appropriately named.
